Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-26 main.Rmd
axis labels fix, caption first line fixed and Di FIXME labels

 The methods are implemented in the open source R package gravitas. The functions for creating granularities and exploring the associated time series are consistent with a tidy workflow( FIXME REF), and the probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk]. 
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\hline
circular granularity && expression  && period length\\
\hline
minute-of-hour & $C_1$ & $= L_1(z \mod 60)$  & $P_1 = $&$60$ \\
minute-of-day & $C_2$ & $= L_2(z \mod 60*24)$  & $P_2=$&$`r 60*24`$\\
hour-of-day & $C_3$ & $= L_3(\lfloor z/60\rfloor\mod 24)$  & $P_3 =$&$24$ \\
hour-of-week & $C_4$ & $= L_4(\lfloor z/60\rfloor\mod 24*7)$  & $P_4=$&$`r 24*7`$\\
day-of-week & $C_5$ & $= L_5(\lfloor z/24*60\rfloor \mod 7)$  & $P_5=$&$7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of circular granularities with bottom granularity minutes. Circular granularity $C_i$ relates two linear granularities one of which groups periodically into the other with regular mapping and $P_i$ represents the period length.}
\label{tab:definitions}
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\hline
quasi-circular granularity && potential period lengths\\
\hline
$Q_1 =$ day-of-month && $P_1 = 31, 30, 29, 28$\\
$Q_2 =$ hour-of-month && $P_2 = 24*31, 24*30, 24*29, 24*28$\\
$Q_3 =$ day-of-year && $P_3 = 366, 365$\\
$Q_4 =$ week-of-month && $P_4 = 5, 4$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of quasi-circular granularities $Q_i$ with potential period lengths $P_i$. These quasi-circular granularities relate two linear granularities one of which groups periodically into the other with irregular mapping and hence can not be expressed through modular arithmetic.}
\label{tab:quasi}
\end{table}

@@

```{r quasi-circular-example, echo = FALSE, out.width = "100%",fig.cap="The time domain distributed as linear granularities $B$, $G'$, $H'$ and quasi-circular granularities $Q(B, G')$ and $Q(B, H')$. Both $Q(B, G')$ and $Q(B, H')$ have repeating patterns unlike $G'$ or $H'$ and the patterns are guided by the distribution of $B$ within $G'$ and $H'$ within a period shown by the hashed arrows."}
knitr::include_graphics("Figs/quasi-circular-ex.png")
```

@@

```{r aperiodic-example, echo=FALSE,out.width = "100%",fig.cap="The time domain distributed as linear aperiodic granularities $B$, $M1$, $M2$, $M$ and aperiodic cyclic granularity $A(B, M)$. $A(B, M)$ have repeating patterns with three categories, two of which correspond to the aperiodic linear granularities $M1$ and $M2$ and other one correponding to no aperiodic granularity."}
knitr::include_graphics("Figs/aperiodic-ex.png")
```

@@

```{r allFig, out.width="100%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Letter value plot of electricity consumption data (a) across month-of-year faceted by quarter-of-year, (b) across weekday/weekend faceted by quarter-of-year and (c) across quarter-of-year faceted by weekday/weekend. Plots (b) and (c) both show harmonies since each quarter consists of both weekdays and weekends. Conversely, weekend and weekday can occur in every quarter. Plot (b) helps to compare weekends and weekdays for each quarter. It can be seen that for every quarter weeekend and weekday consumption are fairly similar except for the second quarter where the letter value D and E behave differently, whereas, Plot (c) helps to compare quarters within weekdays and weekends. For example, the quartile spread of consumption shrinks/lowers from first to fourth quarter, whereas this pattern is not true for weekends. Plot (a) shows a clash since all quarters do not correspond to all months of the year. ", fig.show = 'hold'}

@@

  knitr::kable(smart_harmony,
               format = "latex",
               booktabs = TRUE,
                caption = "Harmonies with two cyclic granularity one placed on facet and the other on x-axis. Out of 30 possible combinations of cyclic granularities, only these 13 are harmony pairs.") %>% 
    #row_spec(0, bold = TRUE) %>% 
    kable_styling()
```

@@

```{r bothcust, fig.cap = "Energy consumption of a single customer shown with different distribution displays, and granularity arrangements. Two granularities are used: hour of the day (I) and  weekday/weekend (II). Plot (a) shows granularity I facetted by granularity II, and plots (b), (c) shows the converse mapping. Plot (a) makes a comparison of usage by workday within each hour of the day using side-by-side boxplots. Generally, on a work day there is more consumption early in the day.  Plots (b) and (c) examine the temporal trend of consumption over the course of a day, separately for the type of day. Plot (b) uses an area quantile to put the emphasis on the time series, for example, the median consumption over time shows prolonged usage in the morning on weekdays. Plot (c) uses a violin plot to place emphasis on distributional differences across hours. It can be seen that the morning use on weekdays is bimodal, some work days there is low usage, which might indicate the person is working from home and also having a late start. (FIXME: make nice axis labels)", out.width="90%"}

# plot for only customer is shown
# cust1 <- smart_meter10 %>%
# filter(customer_id %in% c(10006704)) %>%
#   prob_plot("wknd_wday",
#                             "hour_day",
#                             response = "general_supply_kwh",
#                             plot_type = "quantile",
#             symmetric = FALSE,
#             quantile_prob = c(0.25, 0.5, 0.75, 0.9))
#   scale_y_sqrt() +
#   ggtitle("Energy consumption distribution for customer id: 10006704")
cust2_quantile <- smart_meter10 %>%
  filter(customer_id %in% c(10017936)) %>%
  prob_plot("wknd_wday",
    "hour_day",
    response = "general_supply_kwh",
    plot_type = "quantile",
    symmetric = TRUE,
    quantile_prob = c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99)
  ) +
  scale_y_sqrt() +
  ggtitle(" (b) Area quantile plot faceted by weekend-weekday")  + ylab("Electricity Demand [KWh]")
#
# smart_meter10 %>%
#   filter(customer_id %in% c(10006704)) %>%
#   prob_plot("wknd_wday",
#                             "hour_day",
#                             response = "general_supply_kwh",
#                             plot_type = "violin") +
#   scale_y_sqrt() +
#   ggtitle("Energy consumption distribution for customer id: 10006704")

cust2_violin <- smart_meter10 %>%
  filter(customer_id %in% c(10017936)) %>%
  prob_plot("wknd_wday",
    "hour_day",
    response = "general_supply_kwh",
    plot_type = "violin"
  ) +
  scale_y_sqrt()+
  ggtitle(" (c) Violin plot faceted by weekend-weekday") + ylab("")

smart_meter10 %>%
  filter(customer_id %in% c(10017936)) %>%
  prob_plot("hour_day",
    "wknd_wday",
    response = "general_supply_kwh",
    plot_type = "boxplot"
  ) +
  scale_y_sqrt()+
  ggtitle(" (a) Box plot faceted by hour-of-day")  + ylab("Electricity Demand [KWh]")

ggarrange(cust2_quantile, cust2_violin, nrow = 2)
```

@@

```{r cricex, fig.cap= "(a) Letter value plot shows that there is no upward shift in runs in the second innings like that in the first implying teams are vulnerable to score more in the first innings as they approach the end of the inning. (b) Box plot shows at least one wicket in the previous over leads to downward shift of median run rate and quartile spread in the subsequent over for most overs.", warning=FALSE, message = FALSE, out.width = "90%"}

@@

cricket_tsibble %>%
  filter(batting_team %in% c(
    "Mumbai Indians",
    "Chennai Super Kings"
  )) %>%
  prob_plot("inning",
    "over",
    response = "runs_per_over",
    hierarchy_model,
    plot_type = "lv"
  ) + 
  scale_fill_brewer(palette = "Dark2") + 
  ggtitle("(a) Runs per over across over faceted by inning") +
  theme(legend.position = "right")

@@

# filter(fielding_wckts %in% c(0,1)) %>%
# 
 cricket_data %>%
  filter(over!=1) %>% 
  prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "boxplot",
  symmetric = FALSE,
 quantile_prob = c(0.25, 0.5, 0.75)
) + ggtitle("(b) Runs per over across overs faceted by number of wickets in previous over") + 
  #ylab("runs per over")  +
  xlab("Number of wickets in previous over")
# + geom_boxplot(width = 0.5, aes(colour = lag_field)) 
```

>>>>

 The methods are implemented in the open source R package gravitas. The functions for creating granularities and exploring the associated time series are consistent with a tidy workflow (@Grolemund2018-po), and the probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk]. 
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\hline
circular granularity && expression  && period length\\
\hline
minute-of-hour & $C_1$ & $= L_1(z \mod 60)$  & $P_1 = $&$60$ \\
minute-of-day & $C_2$ & $= L_2(z \mod 60*24)$  & $P_2=$&$`r 60*24`$\\
hour-of-day & $C_3$ & $= L_3(\lfloor z/60\rfloor\mod 24)$  & $P_3 =$&$24$ \\
hour-of-week & $C_4$ & $= L_4(\lfloor z/60\rfloor\mod 24*7)$  & $P_4=$&$`r 24*7`$\\
day-of-week & $C_5$ & $= L_5(\lfloor z/24*60\rfloor \mod 7)$  & $P_5=$&$7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of circular granularities with bottom granularity minutes. Circular granularity $C_i$ relates two linear granularities one of which groups periodically into the other with regular mapping and $P_i$ represents the period length. Circular granularities could be expressed using modular arithmetic due to their regular mapping. }
\label{tab:definitions}
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\hline
quasi-circular granularity && potential period lengths\\
\hline
$Q_1 =$ day-of-month && $P_1 = 31, 30, 29, 28$\\
$Q_2 =$ hour-of-month && $P_2 = 24*31, 24*30, 24*29, 24*28$\\
$Q_3 =$ day-of-year && $P_3 = 366, 365$\\
$Q_4 =$ week-of-month && $P_4 = 5, 4$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of quasi-circular granularities $Q_i$ with potential period lengths $P_i$. These quasi-circular granularities relate two linear granularities one of which groups periodically into the other with irregular mapping and hence can not be expressed through modular arithmetic. }
\label{tab:quasi}
\end{table}

@@

```{r quasi-circular-example, echo = FALSE, out.width = "100%",fig.cap="The time domain distributed as linear granularities $B$ (bottom granularity), $G'$, $H'$ and quasi-circular granularities $Q(B, G')$ and $Q(B, H')$. Both $Q(B, G')$ and $Q(B, H')$ have repeating patterns unlike $G'$ or $H'$ and the patterns are guided by the distribution of $B$ within $G'$ and $H'$ and remains constant within a period shown by the hashed arrows."}
knitr::include_graphics("Figs/quasi-circular-ex.png")
```

@@

```{r aperiodic-example, echo=FALSE,out.width = "100%",fig.cap="The time domain distributed as linear aperiodic granularities $M1$, $M2$, $M$ and aperiodic cyclic granularity $A(B, M)$ relating them with bottom granularity $B$. $A(B, M)$ have repeating patterns with three categories, two of which correspond to the aperiodic linear granularities $M1$ and $M2$ and other one correponding to no aperiodic granularity."}
knitr::include_graphics("Figs/aperiodic-ex.png")
```

@@

```{r allFig, out.width="100%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Distribution of energy consumption displayed through letter value plots. Plot (a) displays consumption across month-of-year faceted by quarter-of-year, (b) across weekday/weekend faceted by quarter-of-year and (c) across quarter-of-year faceted by weekday/weekend. Plots (b) and (c) both show harmonies since each quarter consists of both weekdays and weekends. Conversely, weekend and weekday can occur in every quarter. Plot (b) helps to compare weekends and weekdays for each quarter. It can be seen that for every quarter weeekend and weekday consumption are fairly similar except for the second quarter where the letter values below D and E behave differently, whereas, Plot (c) helps to compare quarters within weekdays and weekends. For example, the quartile spread of consumption shrinks/lowers from first to fourth quarter, whereas this pattern is not true for weekends. Plot (a) shows a clash since all quarters do not correspond to all months of the year. ", fig.show = 'hold'}

@@

  knitr::kable(smart_harmony,
               format = "latex",
               booktabs = TRUE,
                caption = "Harmonies with two cyclic granularity one placed on facet and the other on x-axis along with their number of categories/levels. Out of 30 possible combinations of cyclic granularities, only these 13 are harmony pairs.") %>% 
    #row_spec(0, bold = TRUE) %>% 
    kable_styling()
```

@@

```{r bothcust, fig.cap = "Energy consumption of a single customer shown with different distribution displays, and granularity arrangements. Two granularities are used: hour of the day (I) and  weekday/weekend (II). Plot (a) shows granularity I facetted by granularity II, and plots (b), (c) shows the converse mapping. Plot (a) makes a comparison of usage by workday within each hour of the day using side-by-side boxplots. Generally, on a work day there is more consumption early in the day.  Plots (b) and (c) examine the temporal trend of consumption over the course of a day, separately for the type of day. Plot (b) uses an area quantile to put the emphasis on the time series, for example, the median consumption over time shows prolonged usage in the morning on weekdays. Plot (c) uses a violin plot to place emphasis on distributional differences across hours. It can be seen that the morning use on weekdays is bimodal, some work days there is low usage, which might indicate the person is working from home and also having a late start.", out.width="90%"}

# plot for only customer is shown
# cust1 <- smart_meter10 %>%
# filter(customer_id %in% c(10006704)) %>%
#   prob_plot("wknd_wday",
#                             "hour_day",
#                             response = "general_supply_kwh",
#                             plot_type = "quantile",
#             symmetric = FALSE,
#             quantile_prob = c(0.25, 0.5, 0.75, 0.9))
#   scale_y_sqrt() +
#   ggtitle("Energy consumption distribution for customer id: 10006704")
cust2_quantile <- smart_meter10 %>%
  filter(customer_id %in% c(10017936)) %>%
  prob_plot("wknd_wday",
    "hour_day",
    response = "general_supply_kwh",
    plot_type = "quantile",
    symmetric = TRUE,
    quantile_prob = c(0.01, 0.1, 0.25, 0.5, 0.75, 0.9, 0.99)
  ) +
  scale_y_sqrt() +
  #ggtitle(" (b) Area quantile plot faceted by weekend-weekday") +
  ylab("Electricity Demand [KWh]") + xlab("hour-of-day") + ggtitle("") + ylab("")
#
# smart_meter10 %>%
#   filter(customer_id %in% c(10006704)) %>%
#   prob_plot("wknd_wday",
#                             "hour_day",
#                             response = "general_supply_kwh",
#                             plot_type = "violin") +
#   scale_y_sqrt() +
#   ggtitle("Energy consumption distribution for customer id: 10006704")

cust2_violin <- smart_meter10 %>%
  filter(customer_id %in% c(10017936)) %>%
  prob_plot("wknd_wday",
    "hour_day",
    response = "general_supply_kwh",
    plot_type = "violin"
  ) +
  scale_y_sqrt()+
  #ggtitle(" (c) Violin plot faceted by weekend-weekday") +
  ylab("") + xlab("hours of the day") + ggtitle("") + 
  scale_x_discrete( breaks = seq(0, 23, 5))

cust2_box <- smart_meter10 %>%
  filter(customer_id %in% c(10017936)) %>%
  prob_plot("hour_day",
    "wknd_wday",
    response = "general_supply_kwh",
    plot_type = "boxplot"
  ) +
  scale_y_sqrt() +
  #ggtitle(" (a) Box plot faceted by hour-of-day") +
  xlab("weekday(wdy)/ weekend(wnd)") + ggtitle("") + ylab("") + 
  scale_x_discrete(labels = c('wdy','wnd'))

#ggarrange(cust2_quantile, cust2_violin, nrow = 2)

gg_fig <- ggarrange(cust2_box, 
          ggarrange(cust2_quantile, cust2_violin, nrow = 2, labels = c("(b)", "(c)")),
          ncol = 2, labels = "(a)")
          #label.y = "Electricity Demand [KWh]"\

ggpubr::annotate_figure(gg_fig,
                left = text_grob("Electricity Demand [KWh]",  rot = 90))
```

@@

```{r cricex, fig.cap= "Runs per over shown with different distribution displays, and granularity arrangements. Plot (a) shows letter value plot across overs faceted by innings. There is no upward shift in runs in the second innings like that in the first implying teams are vulnerable to score more in the first innings as they approach the end of the inning. Plot (b) shows box plot across number of wickets in previous over faceted by current over. This indicates that at least one wicket in the previous over leads to lower median run rate and quartile spread in the subsequent over.", warning=FALSE, message = FALSE, out.width = "90%"}

@@

cricket_tsibble %>%
  filter(batting_team %in% c(
    "Mumbai Indians",
    "Chennai Super Kings"
  )) %>%
  mutate(inning = paste0("inning:", inning)) %>% 
  prob_plot("inning",
    "over",
    response = "runs_per_over",
    hierarchy_model,
    plot_type = "lv"
  ) + 
  scale_fill_brewer(palette = "Dark2") + 
  ggtitle("(a) Runs per over across over faceted by inning") +
  theme(legend.position = "right") +
  ggtitle("(a)") +
  ylab("runs per hour") +
  xlab("over-of-inning") 
  # ggplot2::theme(
  #     strip.text = ggplot2::element_text(
  #       size = 7,
  #       margin = ggplot2::margin()
  #     )
  #   ) 
  #

@@

# filter(fielding_wckts %in% c(0,1)) %>%
# 
 cricket_data %>%
  filter(over!=1) %>% 
  prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "boxplot",
  symmetric = FALSE,
 quantile_prob = c(0.25, 0.5, 0.75)
) + ggtitle("(b) Runs per over across overs faceted by number of wickets in previous over") + 
  ylab("runs per over")  +
  xlab("Number of wickets in previous over") +
  ggtitle("(b)")
# + geom_boxplot(width = 0.5, aes(colour = lag_field)) 
```
